Finance and Income Inequality - panel BMA approach\thanks{We thank two anonymous referees, ...}

### 2020-01-05 Paper/paper.tex
update

\section{Introduction}
% \label{sec:intro}

\section{Related literature}
% \label{sec:literature}

@@

\subsection{theory}
\citet{galormoav2004} provide a theoretical framework of thinking about the link between income inequality and financial development. In their model, the income inequality falls due to the decrease of financial friction which allows for physical and human capital accumulation of the poor. That is the extensive margin of financial development at work.

@@

\citet{GreenwoodJovanovic1990} on the contrary present a model where the intensive margin of the financial development is the key force and the benefits of more finance are accrued by the incumbents - primarilly the rich - which implies an inequality increasing effect.

@@

\citet{perotti2007investor} present a theoretical framework how established interests lobby for lower investor protection to prevent entrance of the new competitors. Greater accountability of the politicians is reflected in higher bribe required from teh lobbyist therefore higher accountability implies higher market entry on more competition. After confirming their prediction empirically, they show that the most important factor of accountability is not the formal measure of democratic institutions, but \emph{newspaper readership} which they interpret as broad awareness of policy choices and their outcomes.

@@

\section{Data}
% \label{sec:data}
\begin{figure}
    \caption{Gini Coefficient and Efficiency of Financial Institutions}
    \label{fig:ginifie}
\includegraphics[width=\textwidth, keepaspectratio]{figures/FIEGiniNet}
\end{figure}

>>>>

\section{Introduction}
% \label{sec:intro}
Finance captures the capacity of financial intermediaries and markets to screen investment opportunities, monitor the debtors who were provided funding, as well as pooling and management of risk. With inequality, we focus in the paper on the inequality in the distribution of income. Arguably, the literature discusses other concepts of inequality, e.g. intergenerational persistence of relative income differences or equality of opportunity \citep{demirgucc2009finance}.

\section{Related literature}
% \label{sec:literature}
The research in the area of financial development and income inequality is well established. \citet{demirgucc2009finance}, \citet{claessens2007finance}, and more recently \cite{de2017finance} provide extensive reviews of the topic. The similar theme emerges in all these papers, with the conclusions from both theoretical and empirical contributions provide conflicting prediction about the relationship and well as not unambiguously conclusive empirical results. Tentative conclusions point towards finance tightening the distribution of income. Among other channels, accessing credit allows human capital accumulation and new firm formation and growth equalize economic opportunities. Furthermore, the indirect effect of finance on labour markets through faster economic growth and increased competition diproportionately benefits the low-income individuals.

Conflicting prediction of finance on inequality - intensive vs. extensive margin. Intergenerational persistence might fall with increased economic oportunities while widening the income distribution in every generation.

DIRECT vs. INDIRECT mechanism - access to finance, expanding opportunities of the previsously disadvantaged groups or intensifying the use of financial services by the incumbent individuals and firms. The financial system having second-order effects on inequality by influencing aggregate production and allocation of credit and consequantly the demand for high- vs. low- skilled workers. \citet{braunetal2019} link growth, inequality, and finance in their theoretical model (access to markets inhibits growth).

\citet{delis2014} bank regulations and inequality - nice intro

% Financial development alters how much are economic opportunities depend on the individual skills, family endowments, social status or political connections. Individual depend on financial system to provide loans to start new business, attain education

Both theoretical and empirical studies leave out the issue of importing the financial services from abroad. We include financial globalization from KOF among our control variables.

@@

\subsection{theory}
\textsc{human capital investment} - with imperfect markets restrict the capital accumulation, differences between human (optimal is equal distribution) and physical capital (diminishin returns to physical capital do not materialize at the level of individual ownership of capital because it is not embodies in individuals \cite{demirgucc2009finance}. This implies that equal distribution is not necessarily optimal). \textsc{political consequences} - higher inequality will increase the proportion of population favouring greater redistribution.  \textsc{wage discrimination by specific characteristics, e. g. race}
%
\textsc{investment opportunities, savings}. Minimum investment requirements of fixed costs associated with high-return investments, including entrepreneurship. => higher return to wealthier individuals (e.g. \citet{AtjeJovanovich1993,demel2008qje}).
\textsc{savings behaviour} - if inheritance is a function on income, then rich parents leave larger bequests than the poor and preserve relative differences in incomes. Research indicates that financial development does not alter aggregate savings rates. Is this also valid for more detailed look on poor and rich households? If the effect differs then the distribution of saving is altered by financial development influence the relative income levels across households while keeping the aggregate savings rate unchanged.

\textsc{Summary}

\citet{galormoav2004} provide a theoretical framework of thinking about the link between income inequality and financial development. In their model, the income inequality falls due to the decrease of financial friction which allows for physical and human capital accumulation of the poor. That is the extensive margin of financial development at work.

@@

\citet{GreenwoodJovanovic1990} on the contrary present a model where the intensive margin of the financial development is the key force and the benefits of more finance are accrued by the incumbents - primarily the rich - which implies an inequality increasing effect in the early stages of the development. With tim more agents meet the fixed costs of joining the financial intermediaries and they enjoy higher returns. The efficiency of resource allocation increases, which enhances growth and reduces inequality.

@@

\citet{perotti2007investor} present a theoretical framework how established interests lobby for lower investor protection to prevent entrance of the new competitors. Greater accountability of the politicians leads to a higher bribe required from the lobbyists'. Investor protection strengthens and increases market entry and competition. They examine their prediction in a cross-section and show that better investor protection correlates with larger entry rates and higher firm density in more financially intensive sectors. Furthermore, they show that the most important factor of accountability is not the formal measure of democratic institutions, but \emph{newspaper readership} which they interpret as broad awareness of policy choices and their outcomes. \textsc{size does not matter for entry when investor protection is accounted for}.

@@

\subsection{empirics}
\citet{beck2007finance} show that the development of financial intermediaries is accompanied by faster reductions in income inquality and poverty rates.
\citet{beck2010big} document how bank deregulation in the \ac{us} tightened income distribution by benefiting primarily the bottom 50\% of earners. They attribute the effect to the changes in labour market conditions and relatively higher wages and working hours of the low-skilled workers following the reforms.

\citet{claessens2007finance} argue that althugh the deeper financial systems generally provide better opportunities of access to finance, the relationship is not universal.

The benefits of capital markets liberalization seem to be concetrated to the top of the income distribution. Top quintile of the distribution accrues nearly all of the income growth following the liberalization while the share of middle three quantiles decreases and the bottom remains unaffected \citep{das2003income}.

Diwan (1999) show how labour shares fall follwing a crises only partially recovering in the subsequent years.

\citet{kroszneretal2007} show that financial crises have relative more severe impact on the sectors which depend more on external financing. The consequences of crises on firms relate to institutional environment and materialize through lower production capacity and competition.

\citet{glaeseretal2003} model intitutional subversion by the rich classes and their ability to prevent institutional reforms. Such environment often leads to massive redistribution from the rich to the poor, which is assumed to be less efficient than carefully implemented institutional reforms.

\section{Data}
\cite{roineetal2009}, \citet{nolan2019drivers}, \citet{furceri2019robust} for the determinants of income inequality.

% \label{sec:data}
\begin{figure}
    \caption{Gini Coefficient and Efficiency of Financial Institutions}
    \label{fig:ginifie}
\includegraphics[width=\textwidth, keepaspectratio]{figures/FIEGiniNet}
\end{figure}
